---
title: "Before we Start"
teaching: 10
exercises: 5
questions:
- "What should I remember about R?"
- "Is this an introduction to R?"
objectives:
- "Install latest version of R."
- "Install latest version of RStudio."
- "Navigate the RStudio GUI."
- "Install additional packages using the packages tab."
- "Install additional packages using R code."
keypoints:
- "Use RStudio to write and run R programs."
- "Use `install.packages()` to install packages (libraries)."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("00-")
source("../bin/download_data.R")
```

## This is not an introduction to R

This course assumes a certain level of knowledge about R. We are not going
to cover the basics, and you expected to know how to use the following functionalities in R before starting this course.

- Have R and R-studio installed. 
  Alternatively run everything on rstudio.cloud
- Know how to assign values to variables
- know what a function is, and how we pass input and parameters to it
- be familiar with the %>% operator
- know the basic verbs from dplyr of the tidyverse:
  select
  filter
  mutate
  arrange
  summarise
- be familiar with dataframes
- know how to install and load packages

comments
do math on variables
Get the concept of vectors
subsetting vectors and dataframes
logical tests
NA
read in data from a csv/excel

If any of these topics are unfamiliar, we strongly recommend that you either take one of our introductory courses, read up on the curriculum of one of them, or follow one of the many amazing courses you can find online, before taking this course.


We are going to use the library **`danstat`**. Please install by running this command:
```{r, eval = FALSE}
install.packages("danstat")
```

If that fails... try this:
```{r, eval =F}
install.packages("remotes")
library(remotes)
remotes:install_github("cran/danstat")
```

Three concepts are typically not covered in our introductions.

Factors
dates
lists

So we are starting by taking a look at those.

{% include links.md %}