---
title: "Starting with Data"
teaching: 50
exercises: 30
questions:
- "What else have we forgotten about R?"
- "What is a data.frame?"
- "How can I read a complete csv file into R?"
- "How can I get basic summary information about my dataset?"
- "How can I change the way R treats strings in my dataset?"
- "Why would I want strings to be treated differently?"
- "How are dates represented in R and how can I change the format?"
objectives:
- "Describe what a data frame is."
- "Load external data from a .csv file into a data frame."
- "Summarize the contents of a data frame."
- "Subset and extract values from data frames."
- "Describe the difference between a factor and a string."
- "Convert between strings and factors."
- "Reorder and rename factors."
- "Change how character strings are handled in a data frame."
- "Examine and change date formats."
keypoints:
- "Use read_csv to read tabular data in R."
- "Use factors to represent categorical data in R."
source: Rmd
---

# no no no. You should also know this!

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
source("../bin/download_data.R")
```

```{r, eval = TRUE,  message = FALSE, purl = FALSE}
library(tidyverse)

interviews <- read_csv("../data/SAFI_clean.csv", na = "NULL")
```

## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered (ordinal) or unordered (nominal). Factors
create a structured  relation between the different levels (values) of a
categorical variable, such as days of the week or responses to a question in
a survey. This can make it easier to see how one element relates to the
other elements in a column. While factors look (and often behave) like
character vectors, they are actually treated as integer vectors by `R`. So
you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl = TRUE}
respondent_floor_type <- factor(c("earth", "cement", "cement", "earth"))
```

R will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e`, even though the first element in this vector is
`"earth"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl = FALSE}
levels(respondent_floor_type)

nlevels(respondent_floor_type)
```

Sometimes, the order of the factors does not matter. Other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high").
It may improve your visualization, or it may be required by a particular type of
analysis. Here, one way to reorder our levels in the `respondent_floor_type`
vector would be:

```{r, results = TRUE, purl = FALSE}
respondent_floor_type # current order

respondent_floor_type <- factor(respondent_floor_type, 
                                levels = c("earth", "cement"))

respondent_floor_type # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"cement"`,
`"earth"` is more descriptive than `1`, and `2`. Which one is "earth"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and
need to recode "cement" to "brick".

```{r, results = TRUE, purl = FALSE}
levels(respondent_floor_type)

levels(respondent_floor_type)[2] <- "brick"

levels(respondent_floor_type)

respondent_floor_type
```

So far, your factor is unordered, like a nominal variable. R does not know the
difference between a nominal and an ordinal variable. You make your factor an
ordered factor by using the `ordered=TRUE` option inside your factor function.
Note how the reported levels changed from the unordered factor above to the
ordered version below. Ordered levels use the less than sign `<` to denote
level ranking.

```{r, results = TRUE, purl = FALSE}
respondent_floor_type_ordered <- factor(respondent_floor_type, 
                                        ordered = TRUE)

respondent_floor_type_ordered # after setting as ordered factor
```


### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl = FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to
numbers. Another method is to use the `levels()` function. Compare:

```{r, purl = TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

as.numeric(year_fct)                     # Wrong! And there is no warning...

as.numeric(as.character(year_fct))       # Works...

as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(year_fct)`
* We convert these levels to numeric values using `as.numeric(levels(year_fct))`
* We then access these numeric values using the underlying integers of the
vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `memb_assoc` column from our data frame, convert it into a
factor, and use it to look at the number of interview respondents who were or
were not members of an irrigation association:

```{r factor-plot-default-order, fig.alt = "Yes/no bar graph showing number of individuals who are members of irrigation association", purl = TRUE}
## create a vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see that in
addition to "no"s and "yes"s, there are some respondents for which the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. They do not appear on the plot.
Let's encode them differently so they can counted and visualized in our plot.


```{r factor-plot-reorder, purl = TRUE}
## Let's recreate the vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## replace the missing data with "undetermined"
memb_assoc[is.na(memb_assoc)] <- "undetermined"

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```


## Formatting Dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. As a reminder from earlier in this lesson, the best
practice for dealing with date data is to ensure that each component of your
date is stored as a separate variable. In our dataset, we have a
column `interview_date` which contains information about the
year, month, and day that the interview was conducted. Let's
convert those dates into three separate columns.

```{r, eval = FALSE, purl = FALSE}
str(interviews)
```
We are going to use the package **`lubridate`**, which is included in the
**`tidyverse`** installation but not loaded by default, so we have to load
it explicitly with `library(lubridate)`.

Start by loading the required package:

```{r load-package, message = FALSE, purl = FALSE}
library(lubridate)
```

The lubridate function `ymd()` takes a vector representing year, month, and day,
and converts it to a `Date` vector. `Date` is a class of data recognized by R as
being a date and can be manipulated as such. The argument that the function
requires is flexible, but, as a best practice, is a character vector formatted
as "YYYY-MM-DD".

Let's extract our `interview_date` column and inspect the structure:

```{r, purl = FALSE}
dates <- interviews$interview_date
str(dates)
```

When we imported the data in R, `read_csv()` recognized that this column
contained date information. We can now use the `day()`, `month()` and  `year()`
functions to extract this information from the date, and create new columns in
our data frame to store it:


```{r, purl = FALSE}
interviews$day <- day(dates)
interviews$month <- month(dates)
interviews$year <- year(dates)
interviews
```

Notice the three new columns at the end of our data frame.

In our example above, the `interview_date` column was read in correctly as a
`Date` variable but generally that is not the case. Date columns are often read
in as `character` variables and one can use the `as_date()` function to convert
them to the appropriate `Date/POSIXct`format.

Let's say we have a vector of dates in character format:

```{r, purl = FALSE}
char_dates <- c("7/31/2012", "8/9/2014", "4/30/2016")
str(char_dates)
```

We can convert this vector to dates as :

```{r, purl = FALSE}
as_date(char_dates, format = "%m/%d/%Y")
```

Argument `format` tells the function the order to parse the characters and
identify the month, day and year. The format above is the equivalent of 
mm/dd/yyyy. A wrong format can lead to parsing errors or incorrect results.

For example, observe what happens when we use a lower case y instead of upper
case Y for the year. 
```{r, purl = FALSE}
as_date(char_dates, format = "%m/%d/%y")
```

Here, the `%y` part of the format stands for a two-digit year instead of a
four-digit year, and this leads to parsing errors.

Or in the following example, observe what happens when the month and day
elements of the format are switched.

```{r, purl = FALSE}
as_date(char_dates, format = "%d/%m/%y")
```

Since there is no month numbered 30 or 31, the first and third dates cannot be
parsed.

We can also use functions `ymd()`, `mdy()` or `dmy()` to convert character
variables to date.

```{r, purl = FALSE}
mdy(char_dates)
```

# Wrangling data with dplyr

dplyr is a package that makes wrangling data easier.

We wrangle data when we select, filter and summarise data.


The pipe construct makes it easy to string together different
manipulations of the data:
```{r eval = F}
data %>% filter(some logical test on a column)
```

We select a set of columns by using the select function:

```{r}
interviews %>% select(village, memb_assoc)
```

We select a set of rows by using the filter function:

```{r}
interviews %>% filter(village == "Chirodzo")
```

We make a new column using the mutate function:
```{r}
interviews %>% mutate(new_column_name = no_membrs * 10)
```

We calculate summary statistics by using the summarize function:
```{r}
interviews %>% summarise(avg_membrs = mean(no_membrs))
```

Summary statistics are normally combined with the function group_by:
```{r}
interviews %>% group_by(village) %>% 
  summarise(avg_membrs = mean(no_membrs))
```



{% include links.md %}
